---
title: "04_analysis"
format: html
editor: visual
output-file: results/04_analysis.html
---

# Analysis of gene expressions

To identify genes that might have an impact on type 2 diabetes development, we want to select genes with large expression differences between type 2 diabetes (T2D) and control groups.

### Gene expression differences

We will for each data set use the median of the samples of each gene and save them:

```{r}
# selects all columns that start with "GSM", and calculate median across rows/genes.
# then identifyer (gene name) and the median are selected.

# For type 2 diabetes:
df_T2D_medians <- df_T2D |>
  mutate(median_expression_T2D = rowMedians(as.matrix(across(starts_with("GSM"))))) |>
  select(IDENTIFIER, median_expression_T2D)

# For control group:
df_non_T2D_medians <- df_non_T2D |>
  mutate(median_expression_control = rowMedians(as.matrix(across(starts_with("GSM"))))) |>
  select(IDENTIFIER, median_expression_control)
```

Now we will merge the T2D and control medians that we created above, by identifier with inner join. We also calculate the difference of the expression levels and the absolute difference.

```{r}
expression_comparison <- df_T2D_medians |>
  inner_join(df_non_T2D_medians, by = "IDENTIFIER") |>
  mutate(
    median_difference = median_expression_T2D - median_expression_control, # difference medians
    abs_difference = abs(median_difference) # absolute difference
  ) |>
  arrange(desc(abs_difference)) # arrange by the largest absolute difference first

top_diff_genes <- head(expression_comparison, 20)

print("Top 20 differentially expressed genes (by sample medians):")
print(top_diff_genes)
```

## Visualization of top gene differences

```{r}
library(ggplot2)

# Prepare data for plotting
top_10_plot <- head(expression_comparison, 10) |>
  mutate(
    direction = ifelse(median_difference > 0, "Up in T2D", "Down in T2D"),
    gene_label = IDENTIFIER
  )

# bar plot
ggplot(top_10_plot, aes(x = reorder(gene_label, median_difference), 
                       y = median_difference, 
                       fill = direction)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Genes with Largest Expression Differences",
    subtitle = "T2D vs Control Groups",
    x = "Gene Identifier",
    y = "Expression Difference (T2D - Control)",
    fill = "Expression Direction"
  ) +
  scale_fill_manual(values = c("Up in T2D" = "red", "Down in T2D" = "blue")) +
  theme_minimal()
```
